---
title: "Preliminary Analysis and Results - Linear Mixed Models (OASIS longitudinal data)"
author: "Preethi Ravikumar"
format: revealjs
editor: visual
---


##  Goal  {.smaller}

The goal is to understand how various cognitive and brain volume measures (e.g., MMSE, CDR, eTIV, nWBV) are influenced by factors such as age, sex, education, or group (demented vs. nondemented) while accounting for repeated measures (longitudinal data).
	
## Fitting the Linear Mixed Model: {.smaller}

● Outcome: nWBV (Normalized Whole Brain Volume) – representing structural brain changes over time.

● Primary Predictor: Age to see if aging correlates with changes in brain volume.

● Secondary Predictor: Socioeconomic Status (SES) to observe if socioeconomic factors
impact brain volume over time.

● Fixed Effects: Age and Socioeconomic Status (SES).

● Random Effects: Subject.ID.
```{r, warning=FALSE, echo=T, message=FALSE}
#install.packages("readr")
library(readr)
oasis_data <- read_csv("oasis_longitudinal.csv")
```

## R Script for fitting linear mixed Model:{.smaller}
```{r, warning=FALSE, echo=T, message=FALSE}
#install.packages("lme4")
library(lme4)

oasis_data$gender <- oasis_data$'M/F'
oasis_data$SubjectID <- oasis_data$'Subject ID'

# Fit Linear Mixed Model
model <- lmer(nWBV ~ Age + SES + (1 | SubjectID), data = oasis_data)
summary(model)
```

## Maximum Likelihood {.smaller}
```{r, warning=FALSE, echo=T, message=FALSE}
# Fit the model with ML for model comparison
model_ml <- lmer(nWBV ~ Age + SES + (1 | SubjectID), data = oasis_data, REML = FALSE)
summary(model_ml)
```
## Results for Maximum Likelihood: {.smaller}


Random Effects:

• (Intercept) Variance: 0.0009979 (Standard Deviation: 0.0316)
  This is the variability in nWBV across different subjects. The random intercept accounts for individual differences in baseline nWBV.
  
• Residual Variance: 0.00006953 (Standard Deviation: 0.00834)
  This is the remaining variance in nWBV after accounting for both fixed and random effects.

Fixed Effects:

• Intercept: 0.9952. This is the estimated baseline nWBV when Age and SES are zero. In your model, it likely represents an estimated baseline close to 1.
  
• Age: -0.00348. This negative estimate suggests that nWBV decreases with increasing age. Specifically, for each one-unit increase in Age, nWBV is expected to decrease by about 0.00348 units, holding SES constant.
  
• t value: -15.804 (a high magnitude t-value, suggesting Age is statistically significant)

• SES: 0.00088. This positive coefficient suggests that as SES increases, there’s a very slight increase in nWBV. However, given the low t-value (0.369), SES may not have a statistically significant effect on nWBV in this model.
  
  
Summary Interpretation:

• Age is a significant predictor of nWBV, with a negative effect indicating cognitive decline as age increases.

• SES appears to have little to no significant effect on nWBV based on this model.

## REML {.smaller}

```{r, warning=FALSE, echo=T, message=FALSE}
# Fit the model with REML (default)
model_reml <- lmer(nWBV ~ Age + SES + (1 | SubjectID), data = oasis_data, REML = TRUE)
summary(model_reml)
```
## Results for REML: {.smaller}

Random Effects:

• (Intercept) Variance: 0.001016 (Standard Deviation: 0.03187). The variance estimate for the random intercept has changed slightly from the ML output (from 0.0009979 to 0.001016), reflecting REML’s slightly different handling of variance components.
  
• Residual Variance: 0.0000697 (Standard Deviation: 0.008349). This value is nearly identical to the ML result, indicating minimal change in the residual variance under REML.


Fixed Effects:

• (Intercept): 0.9956. This represents the estimated baseline nWBV when Age and SES are zero. The intercept estimate is nearly unchanged from the ML estimate.
  
• Age: -0.00348. The negative coefficient for Age is consistent with the ML output, indicating a similar estimated decline in nWBV with increasing age.
  
• t value: -15.761 (also very similar to the ML result, suggesting that Age is a highly significant predictor)

• SES: 0.00088. This coefficient remains small and statistically insignificant, indicating little to no effect of SES on nWBV in this model.

## Comparison of the two methods:{.smaller}

• Fixed Effects: The estimates for Age and SES under REML are nearly identical to those under ML, showing consistent results.

   • Age has a significant negative impact on nWBV, indicating cognitive decline with age.
 
   • SES appears to have little to no significant effect on nWBV.


In summary, REML is appropriate for final model interpretation as it provides the better estimates for variance components, while both methods confirm Age as a key predictor of nWBV decline.

