---
title: "Dataset, table1 and visulaizations"
author: "BDP - Dasia M., Bess T., Preethi R."
format:
  revealjs:
    slide-level: 4   # Heading level for slide breaks
    theme: night    # Choose a theme for the slides (e.g., "simple", "night")
    transition: slide # Slide transition (other options: "fade", "convex", etc.)
    incremental: true # Incrementally reveal bullet points
---

## OASIS-Longitudinal MRI Data in Nondemented and Demented Older Adults {.smaller}
source: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2895005/

summary of the chosen dataset:

- The dataset includes MRI scans of 150 subjects, aged 60-96, collected longitudinally over 373 sessions.

- Subjects scanned at least twice, with visits separated by a year or more, using the same MRI scanner.

- Clinical Dementia Rating (CDR) categorized as nondemented or with very mild to mild Alzheimer’s disease.

- 72 were nondemented throughout the study, while 64 were demented at the initial visit and remained so.

- They were right-handed, consisting of 62 men and 88 women, with 3-4 T1-weighted MRI scans per session.


```{r, warning=FALSE, echo=F, message=FALSE, result ='hide'}

#install.packages("tableone")
library(tableone)

#install.packages("readr")
library(readr)

oasis_longitudinal <- read_csv("oasis_longitudinal.csv")

# Load necessary library
library(tableone)

# Recode M/F for easier handling in Table 1 (optional)
oasis_longitudinal$Gender <- ifelse(oasis_longitudinal$'M/F' == "M", "Male", "Female")

# Define the variables to be included in Table 1
vars <- c("Age", "Gender", "Hand", "EDUC", "SES", "MMSE", "CDR", "eTIV", "nWBV", "ASF")

# Define the stratifying variable (Group)
strata <- "Group"

# Create Table 1
table1 <- CreateTableOne(vars = vars, strata = strata, data = oasis_longitudinal, factorVars = c("Gender", "Hand", "CDR"))

# Print the table with p-values
print(table1, showAllLevels = TRUE, smd = TRUE)

```



## Table1
<style type="text/css">
.tg  {border-collapse:collapse;border-spacing:0;}
.tg td{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  overflow:hidden;padding:10px 5px;word-break:normal;}
.tg th{border-color:black;border-style:solid;border-width:1px;font-family:Arial, sans-serif;font-size:14px;
  font-weight:normal;overflow:hidden;padding:10px 5px;word-break:normal;}
.tg .tg-w7ak{border-color:#ffffff;font-family:Georgia, serif !important;text-align:left;vertical-align:top}
</style>
<table class="tg"><thead>
  <tr>
    <th class="tg-w7ak"></th>
    <th class="tg-w7ak">level</th>
    <th class="tg-w7ak">Converted</th>
    <th class="tg-w7ak">Demented</th>
    <th class="tg-w7ak">Nondemented</th>
    <th class="tg-w7ak">p</th>
  </tr></thead>
<tbody>
  <tr>
    <td class="tg-w7ak">n</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">37</td>
    <td class="tg-w7ak">146</td>
    <td class="tg-w7ak">190</td>
    <td class="tg-w7ak"></td>
  </tr>
  <tr>
    <td class="tg-w7ak">Female</td>
    <td class="tg-w7ak">F</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;24 ( 64.9) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;60 ( 41.1) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;129 ( 67.9) </td>
    <td class="tg-w7ak">&lt;0.001</td>
  </tr>
  <tr>
    <td class="tg-w7ak">Male</td>
    <td class="tg-w7ak">M</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;13 ( 35.1) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;86 ( 58.9) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;61 ( 32.1) </td>
    <td class="tg-w7ak"></td>
  </tr>
  <tr>
    <td class="tg-w7ak">Dom Hand</td>
    <td class="tg-w7ak">R</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;37 (100.0) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;146 (100.0) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;190 (100.0) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;NA</td>
  </tr>
  <tr>
    <td class="tg-w7ak">CDR</td>
    <td class="tg-w7ak">0</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;18 ( 48.6) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;0 (&nbsp;&nbsp;0.0) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;188 ( 98.9) </td>
    <td class="tg-w7ak">&lt;0.001</td>
  </tr>
  <tr>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">0.5</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;19 ( 51.4) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;102 ( 69.9) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;2 (&nbsp;&nbsp;1.1) </td>
    <td class="tg-w7ak"></td>
  </tr>
  <tr>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">1</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;0 (&nbsp;&nbsp;0.0) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;41 ( 28.1) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;0 (&nbsp;&nbsp;0.0) </td>
    <td class="tg-w7ak"></td>
  </tr>
  <tr>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">2</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;0 (&nbsp;&nbsp;0.0) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;3 (&nbsp;&nbsp;2.1) </td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;0 (&nbsp;&nbsp;0.0) </td>
    <td class="tg-w7ak"></td>
  </tr>
  <tr>
    <td class="tg-w7ak">Age</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">&nbsp;&nbsp;79.76 (7.43)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;76.26 (6.94)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;77.06 (8.10)</td>
    <td class="tg-w7ak">0.045</td>
  </tr>
  <tr>
    <td class="tg-w7ak">Education</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">&nbsp;&nbsp;15.46 (2.52)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;13.67 (2.90)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;15.14 (2.74)</td>
    <td class="tg-w7ak">&lt;0.001</td>
  </tr>
  <tr>
    <td class="tg-w7ak">SES</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;1.73 (0.96)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;2.77 (1.20)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;2.39 (1.05)</td>
    <td class="tg-w7ak">&lt;0.001</td>
  </tr>
  <tr>
    <td class="tg-w7ak">MMSE</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">&nbsp;&nbsp;28.68 (1.56)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;24.51 (4.50)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;29.23 (0.88)</td>
    <td class="tg-w7ak">&lt;0.001</td>
  </tr>
  <tr>
    <td class="tg-w7ak">eTIV</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">1459.27 (135.43)</td>
    <td class="tg-w7ak">1485.85 (173.77)</td>
    <td class="tg-w7ak">1495.50 (184.89)</td>
    <td class="tg-w7ak">0.51</td>
  </tr>
  <tr>
    <td class="tg-w7ak">nWBV</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;0.72 (0.04)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;0.72 (0.03)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;0.74 (0.04)</td>
    <td class="tg-w7ak">&lt;0.001</td>
  </tr>
  <tr>
    <td class="tg-w7ak">ASF</td>
    <td class="tg-w7ak"></td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;1.21 (0.11)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;1.20 (0.14)</td>
    <td class="tg-w7ak">&nbsp;&nbsp;&nbsp;1.19 (0.14)</td>
    <td class="tg-w7ak">0.683</td>
  </tr>
</tbody></table>


## Data Visualizations

## Age of Participants

```{r, warning=FALSE, echo=F, message=FALSE} 

counts <- table(oasis_longitudinal$Age)
barplot(counts, main="Age of Participants",
   xlab="Age")
```

## Count of Clinical Dementia Rating
```{r, warning=FALSE, echo=F, message=FALSE} 
counts <- table(oasis_longitudinal$CDR)
barplot(counts, main="Count of Clinical Dementia Rating",
   xlab="CDR")
```
## Gender of Participants
```{r, warning=FALSE, echo=F, message=FALSE} 
counts <- table(oasis_longitudinal$`M/F`)
barplot(counts, main="Gender of Participants",
   xlab="Gender")
```
## Socio Economic Status of Participants
```{r, warning=FALSE, echo=F, message=FALSE} 
counts <- table(oasis_longitudinal$SES)
barplot(counts, main="Socio Economic Status of Participants",
   xlab="SES")
```
## Demented Status of Participants
```{r, warning=FALSE, echo=F, message=FALSE} 
counts <- table(oasis_longitudinal$Group)
barplot(counts, main="Demented Status of Participants",
   xlab="Status")
```
## Total number of visits for each Participant
```{r, warning=FALSE, echo=F, message=FALSE} 
counts <- table(oasis_longitudinal$Visit)
barplot(counts, main="Total number of visits for each Participant",
   xlab="Visits")
```
## Estimated Total Intracranial Volume  vs Age
```{r, warning=FALSE, echo=F, message=FALSE} 
#| echo: false
x <- oasis_longitudinal$eTIV
y <- oasis_longitudinal$Age
plot(x, y, main = "Estimated Total Intracranial Volume  vs Age",
     xlab = "eTIV", ylab = "Age",
     pch = 19, frame = FALSE)
abline(lm(y ~ x, data = oasis_longitudinal), col = "blue")
```
## Normalize Whole Brain Volume vs Age
```{r, warning=FALSE, echo=F, message=FALSE} 
#| echo: false
x <- oasis_longitudinal$nWBV
y <- oasis_longitudinal$Age
plot(x, y, main = "Normalize Whole Brain Volume vs Age",
     xlab = "nWBV", ylab = "Age",
     pch = 19, frame = FALSE)
abline(lm(y ~ x, data = oasis_longitudinal), col = "blue")

    
```
## Scatter Plot of eTIV vs Count of Participants
```{r, warning=FALSE, echo=F, message=FALSE} 
library(ggplot2)
eTIV_count <-
  as.data.frame(table(oasis_longitudinal$eTIV))
  colnames(eTIV_count) <- c("eTIV", "Count")
  ggplot(eTIV_count,
         aes(x=as.numeric(eTIV), y=Count)) +
    geom_point(color="blue", size=3) +
    labs(title="Scatter Plot of eTIV vs Count of Participants", x="eTIV", y="Count of Participants") +
    theme_minimal()
```


